# Exploratory Data Analysis

Data Science Specialization - Univeristy John Hopkins
(Coursera.org)

## Week 4 - Project 2

### Introduction to the Project

Fine particulate matter (PM2.5) is an ambient air pollutant for which there is strong evidence that it is harmful to human health. In the United States, the Environmental Protection Agency (EPA) is tasked with setting national ambient air quality standards for fine PM and for tracking the emissions of this pollutant into the atmosphere. Approximatly every 3 years, the EPA releases its database on emissions of PM2.5. This database is known as the National Emissions Inventory (NEI). You can read more information about the NEI at the [http://www.epa.gov/ttn/chief/eiinformation.html](EPA National Emissions Inventory web site).

For each year and for each type of PM source, the NEI records how many tons of PM2.5 were emitted from that source over the course of the entire year. The data that you will use for this assignment are for 1999, 2002, 2005, and 2008.

### Review Criteria

For each question:

1. Does the plot appear to address the question being asked?
2. Is the submitted R code appropriate for construction of the submitted plot?

#### Making and Submitting Plots 

For each plot you should

1. Construct the plot and save it to a PNG file.
2. Create a separate R code file (plot1.R, plot2.R, etc.) that constructs the corresponding plot, i.e. code in plot1.R constructs the plot1.png plot.

Your code file should include **code for reading the data** so that the plot can be fully reproduced.

You must also include the **code that creates the PNG file**. Only include the code for a single plot (i.e. plot1.R should only include code for producing plot1.png)

3. Upload the PNG file on the Assignment submission page
4. Copy and paste the R code from the corresponding R file into the text box at the appropriate point in the peer assessment.

### Data

The data for this assignment are available from the course web site as a single zip file: [https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip](Data for Project 2).

The data file is a .zip file of 29MB.

```{r, message=TRUE, warning=TRUE, include=FALSE}
####################################################################
## Preparation operation
####################################################################

# Load packages
library(magrittr)
library(dplyr)
library(ggplot2)

# !!! Remember to set up directory before download dataset for analysis
# URL to download
fileURL <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"

# Local data filename
dataFileZIP <- "./Dataset_P2.zip"

# Directory for the dataset
dirFile <- "./Dataset_P2"

# If not exist, download the dataset.zip,
if (file.exists(dataFileZIP) == FALSE) {
      download.file(fileURL, destfile = dataFileZIP)
}

# Uncompress data file
if (file.exists(dirFile) == FALSE) {
      unzip(dataFileZIP)
}

NEI <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")

```
PM2.5 Emissions Data (summarySCC_PM25.rds): This file contains a data frame with all of the PM2.5 emissions data for 1999, 2002, 2005, and 2008. For each year, the table contains number of tons of PM2.5 emitted from a specific type of source for the entire year. Here are the first few rows and dimensions of the dataset NEI:

```{r}
head(NEI)

dim(NEI)

```

**The NEI variables are the following:**

* fips: A five-digit number (represented as a string) indicating the U.S. county
* SCC: The name of the source as indicated by a digit string (see source code
* classification table)
* Pollutant: A string indicating the pollutant
* Emissions: Amount of PM2.5 emitted, in tons
* type: The type of source (point, non-point, on-road, or non-road)
* year: The year of emissions recorded

**Source Classification Code Table (Source_Classification_Code.rds):**

This table provides a mapping from the SCC digit strings int he Emissions table to the actual name of the PM2.5 source. The sources are categorized in a few different ways from more general to more specific and you may choose to explore whatever categories you think are most useful. For example, source “10100101” is known as “Ext Comb /Electric Gen /Anthracite Coal /Pulverized Coal”.

The variables names and dimension of the SCC dataset are as follow:

```{r}
names(SCC)

dim(SCC)
```

### Assignment

The overall goal of this assignment is to explore the National Emissions Inventory database and see what it say about fine particulate matter pollution in the United states over the 10-year period 1999–2008. You may use any R package you want to support your analysis.

For the analysis, I used the following packaged:
- dplyr
- magritrr
- ggplot2

You must address the following questions and tasks in your exploratory analysis. For each question/task you will need to make a single plot. Unless specified, you can use any plotting system in R to make your plot.

### Questions

#### Question 1

**Have total emissions from PM2.5 decreased in the United States from 1999 to 2008?**

Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

```{r}
# Use Aggregate() function to compute the sum of the emissions by year
Yearlytotals <- aggregate(Emissions ~ year,NEI, sum)

# Plot the result using base plot system
barplot(Yearlytotals$Emissions,
        names.arg = Yearlytotals$year,
        xlab = "Year",
        ylab = "PM2.5 Emissions (Tons)",
        main = "Total PM2.5 Emissions From All US Sources")
```


As diplayed by the barplot above, the total emissions have globally decreased from 1999 to 2008 in the United States.

#### Question 2

**Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008?**

Use the base plotting system to make a plot answering this question.

```{r}
# Extract the data for hte city of Baltimore
BALT_NEI <- NEI[NEI$fips == "24510", ]

# Use aggregate() function to compute the sum of the yearly emission in the city of Baltimore
BALT_total <- aggregate(Emissions ~ year, BALT_NEI, sum)

# Plot the result using base plot system
barplot(BALT_total$Emissions,
        names.arg = BALT_total$year,
        xlab = "Year",
        ylab = "PM2.5 Emissions (Tons)",
        main = "Total PM2.5 Emissions in Baltimore")
```

From this graph, we can see that the total emissions of PM2.5 particles have decreased from 1999 to 2008 Baltimore City.

#### Question 3

**Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City?**

To answer this questions
ggp <- ggplot(baltimoreNEI,aes(factor(year),Emissions,fill=type)) +
  geom_bar(stat="identity") +
  theme_bw() + guides(fill=FALSE)+
  facet_grid(.~type,scales = "free",space="free") + 
  labs(x="year", y=expression("Total PM"[2.5]*" Emission (Tons)")) + 
  labs(title=expression("PM"[2.5]*" Emissions, Baltimore City 1999-2008 by Source Type"))

**Which have seen increases in emissions from 1999–2008? Use the ggplot2 plotting system to make a plot answer this question.**

#### Question 4

**Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?**

#### Question 5

**How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?**

#### Question 6

**Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?**
