---
title: "cachematrixREADME"
author: "Pierre Baudin"
date: "December 26, 2016"
output: html_document
---

# Peer-graded assignment - Lexical scoping

## makeCacheMatrix script

### Description

**makeCacheMatrix** function creates a special "matrix" object that can cache its inverse
We use _**solve**_ function to calculate the inverse of the matrix

For this assignment, we assume that the matrix supplied is always convertible.
That means the matrix **must be square**.

### Function Definition

```{r}
makeCacheMatrix <- function(x = matrix()) {
      # set m for the cache to NULL
      inv <- NULL
      # STORE a matrix x in mat and flush the cache
      setMatrix <- function(mat) {
            x <<- mat
            inv <<- NULL
      }
      # GET the stored matrix
      getMatrix <- function() x
      # CACHE INVERSE
      CacheInverse <- function(solve) inv <<- solve
      # GET INVERSE from cache
      getMatInverse <- function() inv
      # Return a list of the inner functions
      list(setMatrix = setMatrix, getMatrix = getMatrix,
           CacheInverse = CacheInverse, getMatInverse = getMatInverse)
}
```

### Validation of the function operation

```{r}
# Store a matrix in mat using **makeCacheMatrix** function
# Note the matrix must be square
mat <- makeCacheMatrix(matrix(rnorm(16),4,4))

#list the function in mat
summary(mat)

# Display mat
mat$getMatrix()

```


## cacheSolve script

### Description
**cacheSolve** function computes the inverse of a the special "matrix" object returned by **makeCacheMatrix** function.
If the inverse has already been calculated (and the matrix) has not changed) then cacheSolve retrieves the inverse from the cache.

### Function Definition

```{r}
cacheSolve <- function(x, ...) {
      # Return a matrix that is the inverse of 'x'
      # Get the inverse matrix in cache
      # If there is a value in cache, return cache
      if(!is.null(x$getMatInverse())) {
            message("Cache Inverse is available, display from cache:")
            x$getMatInverse()
      } else {
      # get the matrix, calculate the inverse and store in cache
      inv.mat <- solve(x$getMatrix())
      x$CacheInverse(inv.mat)

      # Return inversed matrix from cache
      inv.mat
      }
}
```

### Validation of the function operation

```{r}
# Calculate the inverse and store in cache
cacheSolve(mat)

# Retrieve inverse from cache
# Note message when cache is available
cacheSolve(mat)
```
## Conclusion

These R functions are able to cache potentially time-consuming computations.
For example, the script can be modified to accomodate other function such as taking the mean of a numeric vector. This is typically a fast operation. However, for a very long vector, it may take too long to compute the mean, especially if it has to be computed repeatedly (e.g. in a loop). If the contents of a vector are not changing, it may make sense to cache the value of the mean so that when we need it again, it can be looked up in the cache rather than recomputed.

In this Programming Assignment we take advantage of the scoping rules of the R language and learned how they can be manipulated to preserve state inside of an R object.
